# Basics {#basics} 

## Some Basic Commands

Before typing commands into the script editor, it might be useful for you to know some of the following basic commands.

1.  To place a comment in the script file, begin the line with #.
2.  To run (execute) a line in the script file, place the cursor on the line, hit Ctrl + Enter (Windows) or Cmd + Enter (macOS).
    -   If we typed the command directly into the console, we only need to hit Enter
3.  To run multiple lines in the script file, select the lines, hit Ctrl + Enter (Windows) or Cmd + Enter (macOS).
4.  To clear the console when it becomes too messy, hit Ctrl + L (Windows) or Cmd + L (macOS).
5.  To request help, type a question mark in front of the command or the package name (e.g., ?cor). Information about the command will appear in the Help tab (lower right pane).

## Typing and Executing Commands

Now, let's actually get R to do stuff for us! Open a script file. Then copy and paste the following R codes into the script file. Run each line (Ctrl/Cmd + Enter) to see what they do. Note that `#` denotes a comment, and therefore it will run as a line of text. Experiment and have fun!

```{r eval=FALSE}
# Assign a single value (e.g., 9) to an object, say x.  
x <- 9  # this means "x gets the value of 9"

# Get the value for x.
x   # remember that R is case-sensitive. If you typed X, you'll get an error message.

# If you want to know what objects are in the workspace (i.e., R's memory), look at the Environments tab or type ls().
ls()

# You may remove an object (e.g., x) from the workspace.
rm(x)  # where rm stands for remove

# If there are too many objects in the workspace, you may remove all objects from the workspace using rm(list=ls()).
rm(list = ls())

# Assign a non-numerical value by putting the value in quotation marks.
y <- "hello!"

# Get the value of y. Notice the value of y is in quotation marks, indicating it is a non-numerical value.
y

# Perform the following mathematical operations in R.
11 + 10  
11 - 10  
11 * 10
11 / 10
11 ^ 10
11 ^ (1/2)
sqrt(11)   # this number should be the same as above line
log(11)    # taking natural log (log base e) 
log10(11)  # taking log base 10
exp(11)    # taking the exponential

# Perform mathematical operations in R with an object (e.g., a). 
a <- 11  # a gets the value of 11
a + 10
a - 10
a * 10
a / 10
a ^ (1/2)
sqrt(a)
log(a)
log10(a)
exp(a)

# Perform mathematical operations with more than one object.
y <- 2  # notice that 2 now replaces the value “hello”. 
a + y
a - y
a * y
a / y
a ^ (1/y)

# You cannot perform mathematical operations with non-numerical objects. 
b <- "1"  # recall, putting things in between quotation marks makes it non-numerical, even if 1 is a number.
a + b  # you get an error here because you cannot perform mathematical operations with non-numerical objects.

# An object can store more than one value, such as a set of numbers or a set of characters. This is known as a vector and can be created using c(). 
num_vector <- c(1, 2, 3, 4, 5)  # numeric vector
fruits <- c("apple", "banana", "orange")  # non-numeric vector

# Get the values stored in the vectors 
num_vector
fruits

# You can do mathematical operations with numerical vectors but not with non-numerical vectors. 
num_vector * 5  # each value in num_vector is multiplied by 5
fruits * 5  # this throws an error that tells you that you cannot use non-numeric values for this operation

```

## Entering Data Directly Into R

Before we can conduct any analysis, we need to give R some data. One way is to key the data directly into R.

Let's say we have three people, Bob, Andrea, and Calvin. We have their ages, the number of children each of them has, and their gender.

| Name   | Age | No. of Children | Gender |
|:-------|:----|:----------------|:-------|
| Bob    | 48  | 1               | Male   |
| Andrea | 47  | 3               | Female |
| Calvin | 49  | 2               | Male   |

First, we create an object called `name`, and assign the three people's names to it.

```{r}

name <- c("Bob", "Andrea", "Calvin")  
# notice the names are between quotation marks
# this  tells R that name is a string (non-numerical) variable

```
If we type `name` into the console, we should get: 

```{r echo = FALSE}

name

```
Next, we create the object `age`, with the three people’s ages. 

```{r}
age <- c(48, 47, 49)  # notice the order of age should match the order of the names
```

If we type `age` into the console, we should get: 

```{r echo = FALSE}
age
```

Now, let's combine the two variables into a data frame, such that each row represents data for one person, and `name` and `age` are in side-by-side columns by using the `data.frame()` function. 

```{r}
data <- data.frame(Name = name, Age = age)  
# Name = name tells R that the variable name should have the header Name. 
# Age = age tells R that the variable age should have the header Age.
```

Type `data` into the console. You should see: 

```{r echo = FALSE}
data
```

Notice that the output has the Name and the Age columns. Name lists the names of the three individuals. Age lists the ages of the three individuals.

The values of Name and Age in the data frame were copied from the original objects. This means that the original objects, name and age, are still in R’s memory. You can see that this is the case from the Environment tab or when you use the ls() function. 

```{r echo = FALSE}
ls()
```
Let’s remove the original variables name and age.

```{r}
rm(name)
rm(age)
```
Now, when you type ls() into the console, you'll see that all is left is the data frame, data. 

```{r echo = FALSE}
ls()
```

If we want to use the variables in the data frame `data`, we need to attach data$ (a dollar sign after the name of the data frame) before the variable name. For example, if we want to know the ages of the three participants:

```{r}
data$Age
# remember that R is case-sensitive. So if you'd typed data$age, you'll get an error.
```
Let’s say we have more variables to add to the data frame `data`: the number of children the person has and the person’s gender. Let’s label the number of children each person has as Children and the gender of each person as Gender.

```{r}
data$Children <- c(1, 3, 2)
data$Gender <- c("male", "female", "male") 
```

Now, when you type `data` into the console, you'll get: 

```{r echo = FALSE}
data
```

Note that we created Children and Gender within the data frame. So if you type Children and Gender without data$, you will get an error message in the console telling you that the object cannot be found. 

Now, let’s say we made a mistake and need to remove Children from the data frame. You can type: 

```{r}
data$Children <- NULL
data # the Children column is now gone
```

Finally, if we want to save the data frame into a .csv file, we use write.csv().

```{r}
write.csv(data, "bobandreacalvin.csv")  
# specify the name and extension of the file (bobandreacalvin.csv). 
```

The .csv file will now be saved as bobandreacalvin.csv within your current directory. If you use an R project, it will be saved in your project directory. Otherwise, it will be in your default directory. If you don't know what the directory is, type getwd() into the console. (Although... If you'd read my instructions earlier, you really should have set the working directory!)
