# Test Yourselves 1 {#tha1}

Before working on this exercise, please read through the [Getting Started](#start) and the [Basics](#basics) sections first, especially if you have no background in R. 

Suggested answers are at the bottom of this page, but please do try the exercise before looking at them. :)  

## Instructions

1. Download and install R.
1. Download and install RStudio.
1. Start an R Project. 
1. Open up a new script file in RStudio.  
1. Suppose we have the starting monthly salary of 10 men and 10 women (see **Data** section below). Enter the data provided into a data frame (call this data frame `dataset`) where one column represents the biological sex of the individual (call this column `BioSex`) and the other column represents the starting monthly salary (call this column `Salary`) of the individual. Make sure you have 20 rows (one for each individual) and 2 columns (one for each variable).
1. Save the data frame into a .csv file. Name this file `data.csv`. Check in the working directory that the .csv file has been created correctly. 

## Data

```{r echo = FALSE}

Man <- c(3530, 4730, 4330, 3560, 4050, 4880, 4190, 3620, 3530, 5070)
Woman <- c(3790, 2720, 3170, 3320, 3190, 2890, 2920, 3390, 3790, 3680)
data <- cbind(Man, Woman)
knitr::kable(data, align = c('c', 'c')) # arrange data into a table, use centre alignment for both columns

```

## Suggested Answers

```{r eval = FALSE}

# Alternative 1
dataset <- data.frame(BioSex = c("man", "man", "man", "man", "man", "man", "man", "man", "man", "man", "woman", "woman", "woman", "woman", "woman", "woman", "woman", "woman", "woman", "woman"), 
                 Salary = c(3530, 4730, 4330, 3560, 4050, 4880, 4190, 3620, 3530, 5070, 3790, 2720, 3170, 3320, 3190, 2890, 2920, 3390, 3790, 3680))

# Alternative 2
# In this alternative, I call the objects sex and money instead so that you can see how they work in data.frame function.

sex <- c(rep("man", 10),
         rep("woman", 10)) 
# tell R to replicate the string "man" 10 times and replicate the string "woman" 10 times. Then combine the results into a list and save it as sex. The first 10 values will be "man" and the next 10 will be "woman".

money <- c(3530, 4730, 4330, 3560, 4050, 4880, 4190, 3620, 3530, 5070, 3790, 2720, 3170, 3320, 3190, 2890, 2920, 3390, 3790, 3680)
# tell R to combine the 20 values into a list and save it as money

dataset <- data.frame(BioSex = sex, 
                      Salary = money) 
# create data frame

# Alternative 3
# Similar to Alternative 2 except that the three steps are combined into one command.

dataset <- data.frame(BioSex = c(rep("man", 10), 
                                 rep("woman", 10)), 
                      Salary = c(3530, 4730, 4330, 3560, 4050, 4880, 4190, 3620, 3530, 5070, 3790, 2720, 3170, 3320, 3190, 2890, 2920, 3390, 3790, 3680))

# Save data file
write.csv(dataset, "data.csv")

```
