# Using tidyverse {#tidyverse} 

_Warning:_ This is a pretty long tutorial. Please set aside more time to go through this before attempting the next assignment!

Now that you have some basics under your belt, let's move on to installing and using packages from `tidyverse`!

The `tidyverse` is a collection of R packages that many data analysts use. You can think of packages as collections of R codes (and more!) that other people have created to make R more powerful or easier to use. In this course, we will be heavily relying on a few of the packages such as `readr`, `dplyr` and `ggplot2` that are in the `tidyverse` collection. Instead of installing each package separately as we normally would, we can simply install `tidyverse` and that would install all the other packages that we want.

## Installing and Loading tidyverse

To install `tidyverse`, use the function `install.packages()`. 

```{r eval = FALSE}

install.packages("tidyverse")

```

Even after we have installed tidyverse, we cannot use it unless we load it. To load it, we need the `library()` function. 

```{r echo = TRUE, message = FALSE, warning = FALSE, error = FALSE}

library(tidyverse)

```

_Note._ You can temporarily load a package without using the `library()` function using the notation `package::function`. This tells R to load the package for a specific chunk of code and not for the entire session. This allows anyone who reads the code to know which package the function comes from. I typically don't do this because it can be quite tiresome to keep typing the name of the package over and over.

However, `package::function` comes in especially handy when we want to help R distinguish between two packages with the same function names. For example, the packages `ggplot2` and `psych` both have a function `alpha()`. R doesn't know which package to choose if you have both packages loaded. So, to specify that you want to use the function from the `psych` package, you should type  `psych::alpha()`. Similarly, if you want to specify that the function comes from the `ggplot2` package, type `ggplot2::alpha()`. 

Now that we have `tidyverse` loaded, let's do stuff with it. Let's start with importing data. 

## Import Data

Usually, R users do not have to enter their data into R directly. Instead, they already have their dataset in various formats, such as .csv, .txt, .sav, .xlsx, etc. So what they need to do is to import the dataset into R. 

In this class, I will use mainly .csv files. To import .csv files, we can use the `read_csv()` function from the `readr` package in `tidyverse`. I have created a hypothetical dataset to demonstrate how to import .csv files. To follow along, please download the dataset here: [SWB.csv](SWB.csv). 

### SWB Dataset

In this hypothetical dataset, 343 participants provided demographic information and responded to the 9-item Materialism Values Survey (MVS) and the 5-item Satisfaction with Life Scale in 2019 (SWLS2019). The same participants were asked to respond to the Satisfaction with Life Scale in 2021 (SWLS2021). The legend for this dataset is as follows: 

| Variable Name   | Variable Label | Value Label | 
|:-------|:----|:----------------|
|pin|participant identification number||
|gender|gender|0 = male, 1 = female|
|marital_status| marital status| 1 = married, 2 = divorced, 3 = widowed|
|have_children|parental status | 0 = no children, 1 = have children|
|mvs_1| My life would be better if I own certain things I don't have.| 1 = strongly disagree, 2 = disagree, 3 = neither disagree nor agree, 4 = agree, 5 = strongly agree|
|mvs_2| The things I own say a lot about how well I'm doing.| 1 = strongly disagree, 2 = disagree, 3 = neither disagree nor agree, 4 = agree, 5 = strongly agree|
|mvs_3| I'd be happier if I could afford to buy more things.| 1 = strongly disagree, 2 = disagree, 3 = neither disagree nor agree, 4 = agree, 5 = strongly agree|
|mvs_4| It bothers me that I can't afford to buy things I'd like.| 1 = strongly disagree, 2 = disagree, 3 = neither disagree nor agree, 4 = agree, 5 = strongly agree|
|mvs_5| Buying things gives me a lot of pleasure.| 1 = strongly disagree, 2 = disagree, 3 = neither disagree nor agree, 4 = agree, 5 = strongly agree|
|mvs_6| I admire people who own expensive homes, cars, clothes. | 1 = strongly disagree, 2 = disagree, 3 = neither disagree nor agree, 4 = agree, 5 = strongly agree|
|mvs_7| I like to own things that impress people.| 1 = strongly disagree, 2 = disagree, 3 = neither disagree nor agree, 4 = agree, 5 = strongly agree|
|mvs_8| I like a lot of luxury in my life.| 1 = strongly disagree, 2 = disagree, 3 = neither disagree nor agree, 4 = agree, 5 = strongly agree|
|mvs_9| I try to keep my life simple, as far as possessions are concerned. (R)| 1 = strongly disagree, 2 = disagree, 3 = neither disagree nor agree, 4 = agree, 5 = strongly agree|
|swls2019_1|In most ways my life is close to my ideal. (2019)|1 = strongly disagree, 2 = disagree, 3 = slightly disagree, 4 = neither disagree nor agree, 5 = slightly agree, 6 = agree, 7 = strongly agree|
|swls2019_2|The conditions of my life are excellent. (2019)|1 = strongly disagree, 2 = disagree, 3 = slightly disagree, 4 = neither disagree nor agree, 5 = slightly agree, 6 = agree, 7 = strongly agree|
|swls2019_3|I am satisfied with my life. (2019)|1 = strongly disagree, 2 = disagree, 3 = slightly disagree, 4 = neither disagree nor agree, 5 = slightly agree, 6 = agree, 7 = strongly agree|
|swls2019_4|So far I have gotten the important things I want in life. (2019) |1 = strongly disagree, 2 = disagree, 3 = slightly disagree, 4 = neither disagree nor agree, 5 = slightly agree, 6 = agree, 7 = strongly agree|
|swls2019_5|In most ways my life is close to my ideal. (2019)|1 = strongly disagree, 2 = disagree, 3 = slightly disagree, 4 = neither disagree nor agree, 5 = slightly agree, 6 = agree, 7 = strongly agree|
|swls2021_1|In most ways my life is close to my ideal. (2021)|1 = strongly disagree, 2 = disagree, 3 = slightly disagree, 4 = neither disagree nor agree, 5 = slightly agree, 6 = agree, 7 = strongly agree|
|swls2021_2|The conditions of my life are excellent. (2021)|1 = strongly disagree, 2 = disagree, 3 = slightly disagree, 4 = neither disagree nor agree, 5 = slightly agree, 6 = agree, 7 = strongly agree|
|swls2021_3|I am satisfied with my life. (2021)|1 = strongly disagree, 2 = disagree, 3 = slightly disagree, 4 = neither disagree nor agree, 5 = slightly agree, 6 = agree, 7 = strongly agree|
|swls2021_4|So far I have gotten the important things I want in life. (2021) |1 = strongly disagree, 2 = disagree, 3 = slightly disagree, 4 = neither disagree nor agree, 5 = slightly agree, 6 = agree, 7 = strongly agree|
|swls2021_5|In most ways my life is close to my ideal. (2021)|1 = strongly disagree, 2 = disagree, 3 = slightly disagree, 4 = neither disagree nor agree, 5 = slightly agree, 6 = agree, 7 = strongly agree|


Now that you have some idea of what the dataset comprises, let's read in the .csv file.


```{r echo = TRUE, message = FALSE, warning = FALSE, error = FALSE}

data <- read_csv("SWB.csv")

# this code is saying...
# read in the .csv file, "SWB.csv" and assign it to the object "data"
# from now on, "data" will refer to the data in this csv file

```

`read_csv()` is not the only function you can use to read in a .csv file. You can also use the built-in R function `read.csv()`. 

```{r eval = FALSE}

data <- read.csv("SWB.csv")

```

If you use `read.csv()`, you don't have to load `tidyverse`. This is because `read_csv()` comes from the `readr` package (and therefore the `tidyverse` collection) whereas `read.csv()` comes pre-installed with R. 

`read.csv()` is supposedly slower than `read_csv()`, which in turn is slower than `data.table`'s `fread()`. For our purposes, though, it really doesn't matter which you use, so long as you can successfully read the .csv file. 

If you have other file types, such as .txt, .sav, .xlsx, you might need different packages. For example, for excel files (.xlsx), you will need the `readxl` package. For SPSS (.sav), you will need the `haven` package or the `foreign` package. For this class, I am very unlikely to use file types aside from .csv, but it is good to be aware of the packages to use if you want to import other file types. 

### Check Imported Dataset

Before conducting any analyses, check that the dataset has been imported correctly. Go to the `Environment` pane (top right pane of RStudio). Click on `data`. The top left pane should now show the imported data in what looks like a spreadsheet. Alternatively, you may type `View(data)` into the console. 

1. **Rows**: The data for each participant is recorded in a single row (e.g., data for Participant 1 is in Row 1)
1. **Columns**: The data for each variable is recorded in a single column. Names of the variables are in the headers for each column

Scroll down to ensure all rows have been imported correctly. There should be 343 rows. Scroll right to ensure all columns have been imported correctly. There should be 23 columns. 

Another way to check the imported dataset is using the `str()` or the `glimpse()` functions. Both give you roughly similar information (e.g., that there are 343 rows and 23 columns, the names of the variables, the values for the first 10 or so participants). Compare them below. 

```{r}

str(data)

```

```{r}

glimpse(data)

```

## Factors

When you look at the output from `str(data)` or `glimpse(data)`, you can see that some values don't seem to make sense on their own. For example, what do 0 and 1 in gender mean? From the legend table earlier, we can see that 0 stands for male and 1 stands for female. But having to consult the table every time we do an analysis is tiresome and also error-prone. So, we use the `factor()` command to tell R what those values mean for categorical variables. 

```{r}

data$gender <- factor(data$gender, levels = c(0, 1), labels = c("male", "female"))
# this code is saying...
# we want to convert the variable gender into a factor
# this variable has two values / levels, 0 and 1 
# label 0 as “male” and 1 as “female”
# notice that male and female are enclosed in quotation marks (because they are non-numerical or string variables).

# Check that gender is now a factor with either glimpse() or str()
# They'll give you the same output
glimpse(data$gender)

```

## Data Wrangling With dplyr

What should we do after importing a dataset? Well, there are many things we can do! We might want to, say, select specific columns to analyse. Or we might want to select only participants who completed all the questions in the survey. Or we might want to create new variables from existing ones. Or we might want to get some descriptive statistics like mean and standard deviation. All of that is made possible with functions in the `dplyr` package in `tidyverse`.

Some of the functions I've found useful in the `dplyr` package are:

| dplyr functions |	Description |
|:---|:---|
|select() | select specific columns |
|filter() |	filter (keep, select) specific rows |
|mutate()	| create new variables / columns |
|summarise()	| summarise values in form of summary statistics |
|group_by()	| apply operations to different groups | 

There are more functions in the dplyr package! Check out the dplyr cheat sheet here: <https://www.rstudio.com/wp-content/uploads/2015/02/data-wrangling-cheatsheet.pdf>.

When you go through the code below, you'll notice that in `dplyr` (and also `tidyverse` in general), the pipe operator `%>%` is used a lot. You can think of it as a symbol for "then". For example `X %>% Y` would be read as: Do X then take the result from that to do Y. Because it gets tiring to type the pipe operator so frequently, there's a shortcut for it: Ctrl/Cmd + Shift + M. 

With that out of the way... Let's go through the five functions!

### Select() 

The `select()` function allows us to select specific columns. This is especially useful if we have many columns to work with and we only want to focus on a few. 

Let's say we only want to select `gender` and the `swls2019_1` to `swls2019_5` columns in the `SWB` dataset. Here's how we would do it: 

```{r}

swls2019_only <- data %>%   # create the subset called "swls2019_only" from "data"
  select(gender, swls2019_1:swls2019_5)  # select gender and the columns swls2019_1 to swls2019_5

```

This is what the first six rows of `swls2019_only` should look like: 

```{r echo = FALSE}

head(swls2019_only)

```

I sometimes also use `select()` to re-arrange the order of the columns. (In `dplyr`, the "correct" function to re-arrange the order of the columns is `arrange()`. `arrange()` allows us to re-arrange rows in addition to columns. But `select()` works too!) 

So, let's say we want the `swls2019_1` to `swls2019_5` columns to come _before_ `gender`. We simply list the `swls2019_1` to `swls2019_5` columns before `gender`.

```{r}

swls2019_genderlast <- data %>%   
  select(swls2019_1:swls2019_5, gender)  

```

This is what the first six rows of `swls2019_genderlast` should look like: 

```{r echo = FALSE}

head(swls2019_genderlast)

```

### Filter()

Maybe we don't want to select specific columns. Instead, we want to select specific kinds of participants to conduct our analyses on (e.g., say from specific treatment groups or maybe only male participants). We use `filter()` to do that. 

New R users sometimes confuse `select()` with `filter()`. Here's how to differentiate the two: `select()` is used to select columns (i.e., variables) whereas `filter()` is used to select rows (i.e., participants). 

So, let's say we want to select only male participants.

``` {r}

male_only <- data %>%	 # create "male_only" subset from "data", 
  filter(gender == "male")  # then filter (keep) only male participants

```
Notice that the double equal sign `==` is used here. In programming languages, the `==` sign is used when we are comparing the left and the right hand side. Here, we're comparing each row of the column `gender` to the word "male". If  that row matches the word "male", we will filter (keep) that row. Otherwise, we will toss it out. 

This is what the first six rows of male_only should look like: 

```{r echo = FALSE}

head(male_only)

```

### Mutate()

Sometimes, we might want to create new variables, say averages or totals. We can do this with `mutate()`. 

Let's say we want to find the average of `swls2019_1` to `swls2019_5` for each participant. We will use the subset we created just now, `swls2019_only` to do this.

```{r}

swls2019_only <- swls2019_only %>% 
  mutate(swls2019_avg = (swls2019_1 + swls2019_2 + swls2019_3 + swls2019_4 + swls2019_5) / 5)
# this code is saying... 
# create a variable called `swls2019_avg` by adding up swls2019_1 to swls2019_5 and then dividing the total by 5
# add this variable to swls2019_only

```

The first 6 rows of `swls2019_only` now looks like this: 

```{r echo=FALSE}

head(swls2019_only)

```

Notice that the `swls2019_avg` is now appended as the last column. 

You can achieve the same results using `rowMeans()`. 

```{r}

swls2019_only <- swls2019_only %>% 
  mutate(swls2019_avg_rowmeans = rowMeans(across(c(swls2019_1:swls2019_5))))
# rowMeans(across(c(swls2019_1:swls2019_5))) tells R that we want to find the mean (average) for each row (hence rowMeans) across the five variables, swls2019_1 to swls2019_5. 

```

Let's compare `swls2019_avg` and `swls2019_avg_rowmeans` for the first 6 rows: 

```{r echo = FALSE}

swls2019_only %>% 
  select(swls2019_avg, swls2019_avg_rowmeans) %>% 
  head(6) %>% 
  print(width = Inf)

```

You should notice that the two columns give you the exact same results.

### Summarise()

The `summarise()` function is used when we want to get summary statistics such as the mean, median, maximum, minimum, etc for a given column in the data frame. 

Let's say we're interested to know the mean (`avg_swls`), minimum (`min_swls`), maximum (`max_swls`), variance (`var_swls`), standard deviation (`sd_swls`), total number of participants (`total`) for `swls2019_avg`. Again, we will use our `swls2019_only` subset.

```{r}

swls2019_only %>% 
  summarise(avg_swls = mean(swls2019_avg), 
            min_swls = min(swls2019_avg), 
            max_swls = max(swls2019_avg), 
            var_swls = var(swls2019_avg), 
            sd_swls = sd(swls2019_avg), 
            total = n())

```

To be honest, though, I think it might be more convenient to get these summary statistics using the `describe()` function from the `psych()` package.

```{r message = FALSE}

# Load package (install the package before loading!)
library(psych)

# Get the most common descriptive statistics
describe(swls2019_only$swls2019_avg)

```

### Group_by()

Sometimes, we might want to apply the same function to different groups of people. For example, we might want to know what the maximum swls score is for males and for females separately. We can use the `group_by()` function to do this.  

Let's say we're interested to know the mean (`avg_swls`), minimum (`min_swls`), maximum (`max_swls`), variance (`var_swls`), standard deviation (`sd_swls`), total number of participants (`total`) for `swls2019_avg` for males and females separately.

```{r}

swls2019_only %>% 
  group_by(gender) %>% 
  summarise(avg_swls = mean(swls2019_avg), 
            min_swls = min(swls2019_avg), 
            max_swls = max(swls2019_avg), 
            var_swls = var(swls2019_avg), 
            sd_swls = sd(swls2019_avg), 
            total = n())

```


Again, I think that the `describeBy()` function from the `psych` package would be much more convenient here. 

```{r}

describeBy(swls2019_only$swls2019_avg, group = swls2019_only$gender)

```

I've shown in some examples above that we can combine different dplyr functions such as `group_by()` and `summarise()` to achieve specific results. I encourage you to play around with the combinations to see what results they lead to!

## Tables and Plots with ggplot2

Okay, so what else can we do with the data? Well, we can create tables and graphs! But why might we do this? Well... Tables and graphs provide us a way to visualize the data. They show us the distribution of the data and allow us to discover any errors or interesting patterns.

Let's start with the frequency (distribution) tables. 

### Frequency Tables

A frequency (distribution) table tells us the number of times each data value occurs for a given variable. 

Let’s say we want to know how many males and how many females participated in the study using the `swls2019_only` subset. The variable is `gender` and the data values we are interested in are `male` and `female`. Let's find out!

Before looking at my code below, try coming up with your own! (HINT: Use the `summarise()` function!) 

```{r}

swls2019_only %>% 
  group_by(gender) %>%  # Group the result by gender
  summarise(freq = n())  # Count how many of each value in the data

```

The output tells us 168 males and 175 females participated in the study. 

The number of times a data value occurs for a given variable is called the absolute frequency. If we want the relative frequency (i.e., the number of times a data value occurs relative to the total number of observations for a given variable), then we divide the absolute frequency by the total number of values.

Again, try this on your own before looking at my code. (HINT: Use the `summarise()` function and then the `mutate()` function!)

```{r}

swls2019_only %>% 
  group_by(gender) %>%   # Group the result by gender
  summarise(freq = n()) %>%  # Count how many of each value in the data 
  mutate (total = sum(freq),  # Count the total number of observations 
          rel_freq = freq / total)  # Take the frequency of each value / total number of observations

```

The output (rel_freq) tells us that slightly less than half the sample are males (0.49) and slightly more than half are females (0.51). 

Relative frequency is a proportion and may be difficult for some people to interpret. For easier interpretation, we can convert the proportions into percentages by taking 100%. 

Try converting the proportions into percentages on your own before looking at my code below!

```{r}

swls2019_only %>% 
  group_by(gender) %>%   
  summarise(freq = n()) %>%   
  mutate (total = sum(freq), 
          rel_freq = freq / total, 
          percentage = rel_freq * 100)  

```


### Bar Plots with ggplot2

Instead of using a frequency table, maybe you want a bar chart or bar plot to visually represent the frequency distribution instead. (Note that bar charts or bar plots are used for categorical variables only.)

We will use `ggplot2`, a package developed for data visualization, which is part of the `tidyverse` collection.  Every plot in `ggplot2` is plotted with three components: data, aesthetics, and geometry.

* Data: specifies the data frame
* Aesthetics: specifies the variables that we want to plot as well as aesthetics, like colour
* Geometry: specifies the type of plot & other modifications

These three components apply when we make any plot in ggplot2. Let's make a bar chart with `ggplot2` with the code below. 

``` {r}

ggplot(data = swls2019_only, aes(x = gender)) +  # dataset used is swls2019_only, x (horizontal axis) is gender
  geom_bar(stat = "count")  # we want a bar plot where the statistic is the count of the values in x 

```

Here's an alternative that will give you the exact same output.

``` {r eval = FALSE}

# Alternative
swls2019_only %>%  # dataset used is swls2019_only
  ggplot(.) +  # R takes swls2019_only in the previous line and places it in the . on this line to tell ggplot that dataset used is swls2019_only
  geom_bar(aes(x = gender), stat="count")  # we want a bar plot where x (horizontal axis) is gender and the statistic is the count of values in x

```
Notice that the `aes()` command can be either in the ggplot line or the geom function line. Where you type it doesn't affect the results at all. (Try this out yourself to confirm!) 

I personally prefer the alternative because it's easier for me to read, but it doesn't matter to me which you use!

### Histograms with ggplot2

We create bar plots for categorical variables (such as gender). For continuous variables which can take on infinite values (such as time), we create histograms. 

A histogram is a visual representation of the frequency table for a given continuous variable (i.e., it shows us the frequency distribution, or the frequency of scores for a given continuous variable).

Let's use `ggplot2` to make a histogram for `swls2019_avg` with a very basic code.

``` {r}

ggplot(data = swls2019_only, aes(x = swls2019_avg)) + 
  geom_histogram()  # we want a histogram, so geom_histogram() is used

```  

As with the bar chart, you could do this instead and the results would be the same: 

``` {r eval = FALSE}

# Alternative
swls2019_only %>% 
  ggplot(.) + 
  geom_histogram(aes(x = swls2019_avg))

```

Now, the histogram that you got is okay, but kinda basic... Maybe you want to customise it. For example, you might want to add a specific label for the y-axis or the x-axis... Or maybe you want to change the colour of the graph! You can customise all of that! Here are some ideas. 

``` {r}

# Assign color “grey” to the object barfill (which we will use below to change the colour of the bars).
barfill <- "red" 

# Assign color “black” to the object barlines (which we will use below to change the colour of the barlines).
barlines <- "black" 

# Specify the data frame and the variables to plot
ggplot(swls2019_only, aes(x = swls2019_avg)) + 
# Specify the type of plot and binwidth
  geom_histogram(binwidth = 0.5, 
# Specify the border of the bars
                   color = barlines, 
# Specify the color of the bar
                   fill = barfill) + 
# Label the x-axis
  scale_x_continuous(name = "2019 Satisfaction With Life Scores", 
# Force the graph to start at x = 0
                     expand = c(0, 0), 
# Force the graph to end at x = 7
                     limits = c(0, 7)) + 
# Label the y-axis
  scale_y_continuous(name = "Frequency", 
# Force the graph to start at y = 0
                     expand = c(0, 0), 
# No maximum limit to y (hence, NA)
                     limits = c(NA, NA)) + 
# Give the histogram a title.
    ggtitle("Histogram of SWLS2019")
```  

Note that when you set limits on the graph, it throws up a warning: *Warning: Removed 2 rows containing missing values (`geom_bar()`).* This is a known issue, but is apparently too complex to resolve (see: https://github.com/tidyverse/ggplot2/issues/4083). You may safely ignore it if your limits are beyond the minimum and maximum values of your axes. For example, suppose your minimum value for the x-axis is 1 and your maximum value is 7. If you set the limits to 0 and 8, it should comprise all values in your dataset. You can find the minimum and maximum values using the `summarise()` function, or simply use `max()` or `min()`.  

### Scatterplot with ggplot2

So far, we've been focusing on single variables (e.g., `gender` and `swls2019_avg`). What if we're interested in the relationship between two continuous variables? Here is where we might want to plot a scatterplot!

A scatterplot is a graph of pairs of values for each subject or individual. Specifically, each individual provides two observations, one for each variable. One of the variables will be plotted along the x-axis and the other along the y-axis. 

So, suppose we're interested in plotting the relationship between two items from the SWLS2019 dataset: `swls2019_1` and `swls2019_2`. Why might we be interested in this? Well, the reasoning goes something like this: Both items supposedly measure satisfaction with life. So, if a participant scores higher on one item than another participant, they should also score higher on the other item. To confirm this, we can plot a scatterplot. 

Here's the basic code for a scatterplot in `ggplot2`:

``` {r}

swls2019_only %>% 
  ggplot(.) + 
  geom_point(aes(x = swls2019_1, y = swls2019_2)) +  # A scatterplot is a plot of points on a graph. So we use geom_point.
  labs(x = "swls2019_1: In most ways my life is close to my ideal",
       y = "swls2019_2: The conditions of my life are excellent", 
       title = "As swls2019_1 increases, swls2019_2 also increases")  # labs() is another way to add labels. You can also add a title using the labs() function, which by default is left-aligned. 

```

From the scatterplot, we see that indeed, as `swls2019_1` increases, `swls2019_2` increases. 

Like the histogram, there are many different customisations beyond the basics I've shown you here. You can refer to this website for more: https://t-redactyl.io/blog/2016/02/creating-plots-in-r-using-ggplot2-part-5-scatterplots.html.

### Last Words

I've only just scratched the surface of plots in this tutorial. For more, visit here: https://t-redactyl.io/tag/r-graphing-tutorials.html. Or here: https://r4ds.had.co.nz/data-visualisation.html. And here's a link to the ggplot2 cheatsheet: https://posit.co/wp-content/uploads/2022/10/data-visualization-1.pdf. 

Now that you've gotten to this point, I think you're ready to do the next assignment! Proceed to the [next section](#tha2)!```````````